---
title: Territorial Rufous!
draft: true
engine: knitr
date: 2023-08-21
author: Cait Harrigan
image: thumbnail.jpng
toc: true
categories:
  - birds
  - code
  - data analysis
  - deep-dive
format:
  html:
      code-fold: true
execute:
  warning: false
  message: false
---

Reading time: `r ifelse(file.size("index.qmd")/2000 < 1, '<1 minute', paste0(round(file.size("index.qmd")/2000), ' minutes'))`

::: {.callout-note icon=false}
# Series Part 4

This is the second post in this deep-dive series.  
You can start from the beginning with [Part 1](https://galiwatch.ca/posts/hummer-data-cleaning-170323).
:::


# bird

::: callout-note
This is the second post in this deep-dive series. See here for [part ii](https://galiwatch.ca/posts/hummer-behaviour-170323/)
:::

## Setup

As previously covered in part i, we're looking at the hummingbird detection data from 2021. We did some preliminary data exploration, and saw some interesting patterns in the visit times of different birds.


## Continuation of hummer-watch data analysis!

## Number of birds in an image for each class

Rufous males can be very territorial. We'd like to know whether we can see evidence of the "chasing away" behaviour.

If this is the case, then the disappearance of the male rufous should lead to an increase in the *absolute number* of Annas who come (but might not affect the number of female rufous. And the later disappearance of the female rufous should also lead to an increase in the number of Annas.

Because it's hard to see this on a minute-to-minute basis, we'll summarize the number of Anna visits by day.

```{r, eval=FALSE}
bird_c <- merge(bird_scaled, weather, all.x=T)
# number of birds per day 

bird_c %>% 
  group_by(Date, Species, Sex) %>%
  summarise(n=n(), `Temp C` = max(`Temp C`), .groups='keep') %>%
  ggplot() + 
  aes(x = Date, y = n, fill = Sex) + 
  geom_bar(stat = 'identity') +
  facet_wrap(~Species) + 
  labs(y = 'Number of bird visits', title = 'Most Rufous visit in Summer, Female Annas visit in Fall,\nMale Annas are year-round')

# Not really what happened

bird_c %>% 
  group_by(Date, Species, Sex) %>%
  summarise(n=n(), `Temp C` = max(`Temp C`), .groups='keep') %>%
  ggplot() + 
  aes(x = n, y = `Temp C`, colour = Sex) + 
  geom_point() + 
  geom_smooth(method = 'lm', formula = 'y~x', se = F, linetype='dashed') +
  facet_wrap(~Species, scales = 'free_x') + 
  labs(x = 'Number of bird visits', title = 'Everyone except Female Annas prefers lower temperatures')

```
Bird co-occurance: how many share an image with another?
For each month, does the 2+ mixture match the single bird mixture

Rufouses probably chase off annas, and maybe the females more so. Seen in the seasonal trend, but this might be migration. But also seen in the difference between the ratio of membership in the single vs. multi occupancy images. 
See the every month the rufouses are around. Never see more than 3 annas in one image. 

This is also apparent on an hourly level!

```{r}
bird_scaled %>%
  mutate(rufous_migrated = (Date >= ymd('2021-07-31'))) %>%
  group_by(Date, hhmm, Species, Sex, rufous_migrated) %>%
  summarise(n=n()) %>%
  ggplot(aes(x=factor(n), fill=Species)) + 
  geom_bar() +
  #facet_wrap(~month(Date, label = T))
  facet_wrap(~rufous_migrated) + 
  theme_pubr()


bird_scaled %>%
  mutate(rufous_migrated = (Date >= ymd('2021-07-31'))) %>%
  filter(Species=='Annas') %>%
  group_by(Date, Hour, rufous_migrated, Sex) %>%
  summarise(n=n()) %>%
  ggplot(aes(x=factor(n), fill=Sex)) + 
  geom_bar() +
  #facet_wrap(~month(Date, label = T))
  facet_wrap(Sex~rufous_migrated) + 
  theme_pubr()

bird_scaled %>%
  group_by(Date, Hour, Species, Sex) %>%
  summarise(n=n()) %>%
  ggplot(aes(x=factor(n), fill=Species)) + 
  geom_bar(position='fill') +
  #facet_wrap(~month(Date, label = T))
  facet_wrap(month(Date, label = T)~Sex) + 
  theme_pubr()
  
```

fit a poisson model to each and evaluate if rate paranmeter is different!

```{r, eval=T}

# total number of birds per day
bird %>%
  filter(Species=='Annas') %>%
  mutate('Rufous Males present' = Date < ymd('2021-07-31')) %>%
  group_by(Date,`Rufous Males present`, Sex, .groups='keep') %>%
  summarize(n=n()) %>%
  ggplot(aes(x=`Rufous Males present`, y = n)) +
  geom_rect(aes(fill = Sex),xmin = -Inf,xmax = Inf, ymin = -Inf,ymax = Inf, show.legend = F) + 
  geom_boxplot() + 
  facet_wrap("~Sex") + 
  stat_compare_means(method = "t.test", label.y = 180, label.x = 1.3) + 
  geom_bracket(xmin = 1, xmax = 2, label = '', y.position = c(170)) +
  theme_pubr() + 
  labs(title='Rufous Males are better at scaring off Female Annas than Male Annas') + 
  scale_fill_manual(values = c('chartreuse1', 'chartreuse4'))

# average number of birds per image per day
bird %>%
  filter(Species=='Annas') %>%
  mutate('Rufous Males present' = Date < ymd('2021-07-31')) %>%
  
  group_by(Date,`Rufous Males present`, Timestamp, Sex, .groups='keep') %>%
  summarize(n=n()) %>%
  group_by(Date, `Rufous Males present`, Sex) %>%
  summarize(n=mean(n), .groups = 'keep') %>%
  
  ggplot(aes(x=`Rufous Males present`, y = n)) +
  geom_rect(aes(fill = Sex),xmin = -Inf,xmax = Inf, ymin = -Inf,ymax = Inf, show.legend = F) + 
  geom_boxplot() + 
  facet_wrap("~Sex") + 
  stat_compare_means(method = "t.test", label.y = 1.41, label.x = 1.3) + 
  geom_bracket(xmin = 1, xmax = 2, label = '', y.position = 1.4) +
  theme_pubr() + 
  labs(title='Rufous Males are better at scaring off Female Annas than Male Annas') + 
  scale_fill_manual(values = c('chartreuse1', 'chartreuse4'))


bird %>%
  group_by(Date, Timestamp, Species, .groups='keep') %>%
  summarize(n=n()) %>%
  #group_by(Date, Species) %>%
  #summarize(n=mean(n), .groups = 'keep') %>%
  ggplot(aes(x=factor(Date), y = n, colour=Species)) + 
  geom_point()+
  geom_smooth(aes(colour =Species), method='lm', formula='y~poly(x,4)')

bird %>%
  #filter(Species=='Annas') %>%
  mutate('Rufous Males present' = Date < ymd('2021-07-31')) %>%
  group_by(Date, `Rufous Males present`, Species, Sex) %>%
  summarize(n=n()) %>%
  ggplot(aes(x=`Rufous Males present`, y = n)) +
  geom_boxplot() +
  facet_grid(rows =vars(Species), cols = vars(Sex), scales='free')



```

```{r, eval=T}
bird  %>%group_by(Timestamp, Species, Sex) %>% summarize(n=n()) %>% ggplot() +aes(x = Timestamp, y=n, color = Species, fill=Species) + geom_bar(stat='identity') + facet_wrap('~Sex') +scale_color_manual(values=c('chartreuse2', 'chocolate3'))+scale_fill_manual(values=c('chartreuse2', 'chocolate3'))
```

```{r eval=T}
bird %>%
  group_by(Date, Month = factor(month(Timestamp, label=T)), Hour= hour(Timestamp), Species, Sex) %>%
  summarise(n=n(), .groups='keep') %>%
  group_by(Month, Hour, Species, Sex,.groups='keep') %>%
  summarize(mean_n = mean(n)) %>%
  ggplot(aes(x=Hour, y = mean_n, colour = Month)) + 
  geom_point() + geom_line() + 
  facet_grid(rows = vars(Species), cols = vars(Sex), scales = 'free')+ 
  labs(y='average number of IDs per day')

```

```{r eval=T}
bird %>%
  mutate(Hour = hour(Timestamp), Month = factor(month(Timestamp, label=T))) %>%
  group_by(Month, Hour, Species, Sex) %>%
  summarise(n=n()) %>%
  ggplot(aes(x = Hour, y = Month, fill=Month, color =Month)) +
  ggridges::geom_density_ridges(alpha= 0.2) + 
  facet_grid(rows = vars(Species), cols = vars(Sex))


bird %>%
  mutate(Hour = hour(Timestamp), Month = factor(month(Timestamp, label=T))) %>%
  ggplot() +
  aes(x=Hour, fill=Month) +
  geom_bar() + 
  theme_pubr() + theme(legend.position = 'right') +
  facet_wrap('~Species', scale='free') +
  labs(x = 'Hour of Day', y = 'Count', title = 'Daylight changes over the months!')
```



